SUPPLEMENTARY:\\Correcting for bias in high-throughput sequencing data

### 2011-08-23 seqbias-paper.tex
parameter sensitivity explored in the suplement

\begin{figure}
\centerline{\includegraphics[width=0.35\textwidth]{overview.eps}}
\caption{An overview of the approach taken: (a) foreground sequences are sampled
from the regions surrounding the starts of mapped reads, (b) background
sequences are sampled by randomly offsetting foreground positions, (c) a
Bayesian network is classifier is trained to discriminate between the set of
sampled foreground and background sequences, (d) and the classifier is evaluated
at each position within a locus, predicting bias.}
\label{fig:overview}
\end{figure}

@@

The model is trained on $n$ sequences, one half labeled as foreground, the other
background, sampled from the reference genome. To obtain the foreground
sequences, we take sequences surrounding (extending 20 nt to either side, by
default) the starts positions of $n/2$ aligned reads. We wish to avoid
overfitting to any short locus with a large abundance of reads, yet we must also
capture the bias introduced by, for exapmle, PCR amplification. We therefore
choose the $n/2$ unique reads with the highest number of duplicates, yet count
each read only once, ignoring duplicates

>>>>

\begin{figure}
\centerline{\includegraphics[width=0.35\textwidth]{overview.eps}}
\caption{An overview of the approach taken: (a) foreground sequences are sampled
from the regions surrounding the starts of mapped reads, (b) background
sequences are sampled by randomly offsetting foreground positions, (c) a
Bayesian network is trained to discriminate between the set of
sampled foreground and background sequences, (d) and the model is evaluated
at each position within a locus, predicting bias.}
\label{fig:overview}
\end{figure}

@@

The model is trained on $n$ sequences, one half labeled as foreground, the other
background, sampled from the reference genome. To obtain the foreground
sequences, we take sequences surrounding (extending 20 nt to either side, by
default) the starts positions of $n/2$ aligned reads. We wish to avoid
overfitting to any short locus with a large abundance of reads, yet we must also
capture the bias introduced by, for example, PCR amplification. We therefore
choose the $n/2$ unique reads with the highest number of duplicates, yet count
each read only once, ignoring duplicates

### 2011-08-23 seqbias-supplement.tex
parameter sensitivity explored in the suplement

>>>>

\section{Sensitivity of Parameters}

The performance of our method is dependent on two parameters: the standard
deviation at which background sequences are sampled, and the degree to which
model complexity in penalized. We have also introduced parameters limiting the
number of parents a node may have ($p_{\text{max}}$) as well as the distance
that an edge may traverse ($d_{\text{max}}$), but these exist only to control
the amount of CPU time used and have a much simpler interpretation: bigger is
better, but slower.

Here we retrain the model on 50,000 reads from the Mortazavi data
set, varying these first two parameters to assess the sensitivity of the
observed results to their values.

\begin{center}
\includegraphics[width=0.8\textwidth]{fig/pois-sampstd.pdf}
\end{center}

The above plot show the standard deviation used to draw background samples
varied from 1 to 10,000. Apparent from this plot is that, while the optimal
choice lies somewhere between 10 and 100, the model performs competitively for
any reasonable value.

\begin{center}
\includegraphics[width=0.8\textwidth]{fig/pois-comppen.pdf}
\end{center}

Here we multiply the complexity penalty term of the BIC by constants varying
from $10^{-6}$ to $10^{3}$. (Recall that the BIC is $2\ell - m \log n$ where
$\ell$ is the log-likelihood, $m$ is the number of parameters needed to specify
the model, and $n$ is the number of training examples. Here we compute $2 \ell -
c m \log n$, varying $c$.)

It is clear from the above figure, that a very severe complexity penalty will
result in a poor model. Here, for constants larger than 100, an empty model is
trained, resulting in the goodness of fit of 0. Conversely, if this constant is
very small, a overly-dense model will be trained.  In this data, for constants
less than 0.01, the maximally dense model is trained (given the restraints on
in-degree and edge distance). This result is a sub-optimal but entirely adequate
solution.
